---
title: 'Lab: Class One Survey Analysis'
output:
  html_document: default
---


### **INSTRUCTIONS**
#### **Complete all tasks 1-7 using reproducible coding practices (annotation of all code). Use R syntax to solve all problems (i.e. do not manually change values/variables in datasets or check them manually to make sure things are correct except as a double or triple check). Please cut and paste each problem before each R markdown chunk and annotate your code within the chunks.**


##### 1. Import class one survey data supplied by the course (https://github.com/kijohnson/Advanced-Data-Analysis) (The dataset is called 'Class 1 Survey Fall 2026.csv'), calling the R dataframe that you create *C1survey*.

```{r task1, message=FALSE}
# Open the readr library for read_csv(), as demonstrated in the Starbucks demo.
library(readr)

# Open the dplyr library for data management functions used later in the lab.
library(dplyr)

# Import the unmodified course CSV from this repository and call it C1survey.
C1survey <-
  read_csv(
    "Class 1 Survey Fall 2026.csv"
  )
```


##### 2. Determine: **a.** the number of observations (i.e. the number of people who filled out the survey) and **b.** the number of variables in the dataframe.

```{r task2}
# Use dim(), as in the Starbucks demo, to display the number of rows and columns.
# The first value is the number of observations and the second is the number of variables.
dim(C1survey)

# Answer: There are 31 observations and 27 variables in the current course CSV.
```


##### 3. **a.** Rename the column variables to something shorter and that is descriptive of what the variable is about (for example *like_dogs* for the 'Do you like dogs?' question variable) and **b.** Display that they are renamed. **Hints**: You can use either the rename_with function or the names function. The list of names for time is:

```{r task3}
# Use names(), one of the methods suggested in the question, to rename all
# variables in the same order as the columns in the original dataframe.
names(C1survey) <- c(
  "id", "like_cats", "like_dogs", "have_desert", "slogan", "fav_day",
  "larkORowl", "fav_food", "fav_drink", "fav_season", "fav_month", "hobby",
  "program", "specialization", "stat_software", "R_exp", "coding_comfort",
  "coding_length", "top_three", "public_health_interest", "fav_num", "bday",
  "bmonth", "country", "state", "city", "highest_educ_level"
)

# Use colnames(), as in the Starbucks demo, to display the renamed variables.
colnames(C1survey)
```


##### 4. Write code to determine and display the number of factor, integer, numerical, and character variables there are in the *C1survey* dataset. **Suggestion**: Try to ask ChatGPT for a solution if you are unsure. There is a really short answer to this but we want to see what students come up with!

```{r task4}
# Apply the class() function to each variable using sapply(), and then use
# table() to count how many variables belong to each class. These functions
# were all demonstrated in the Starbucks demo.
table(sapply(C1survey, class))

# Answer: There are 23 character variables and 4 numerical variables.
# No factor or integer classes appear in the table, so there are 0 factor
# variables and 0 integer variables.
```


##### 5. **a.** Using code, check the *bday* and *bmonth* variables for any unusual or missing values (**hint**: use table function). If you find any, **b.** describe what you will do with the unusual values in a comment before or immediately following the code that does it, and **c.** after you have cleaned up any unusual values, find the median bday and month. *Note: please make sure that your bday and month variables are numeric before doing statistics.*

```{r task5}
# Use table(), as in the Starbucks demo, to inspect all bday and bmonth values.
# useNA = "always" makes the missing values visible in the tables.
table(C1survey$bday, useNA = "always")
table(C1survey$bmonth, useNA = "always")

# The unusual bday values are "30-Nov" and "29-Nov". Use mutate() and
# if_else(), as demonstrated in the Starbucks demo, to retain their day portions.
C1survey <- C1survey %>%
  mutate(bday = if_else(
    bday == "30-Nov", "30",
    if_else(bday == "29-Nov", "29", bday)
  ))

# The unusual bmonth values are May, March, November, and February (which may
# retain a trailing space depending on how the CSV reader handles whitespace).
# Use nested if_else() functions to change the written months to their numeric
# month codes. The genuine missing value is left as NA.
C1survey <- C1survey %>%
  mutate(
    bmonth = if_else(
      bmonth == "May", "5",
      if_else(
        bmonth == "March", "3",
        if_else(
          bmonth == "November", "11",
          if_else(
            bmonth == "February " | bmonth == "February", "2", bmonth
          )
        )
      )
    )
  )

# Convert the cleaned character variables to numeric using as.numeric(), as
# demonstrated with the calories variable in the Starbucks demo.
C1survey$bday <- as.numeric(C1survey$bday)
C1survey$bmonth <- as.numeric(C1survey$bmonth)

# Check the variable classes after conversion.
class(C1survey$bday)
class(C1survey$bmonth)

# Display the cleaned values as a final check.
table(C1survey$bday, useNA = "always")
table(C1survey$bmonth, useNA = "always")

# Calculate the medians. na.rm = TRUE excludes the genuine missing response.
median(C1survey$bday, na.rm = TRUE)
median(C1survey$bmonth, na.rm = TRUE)

# Answer: The median birth day is 17.5 and the median birth month is 8.5.
```


##### 6. **a.** Create a new variable called *bseason* that gives the season according to Northern Meteorological season in which respondents were born (winter = 12/1 to 2/29, spring = 3/1 to 5/31, summer = 6/1 to 8/31, fall = 9/1 to 11/30). **b.** Using the table function, print a table of Seasons in the columns and bmonths in the rows that allows you to check that the coding is correct. **c.** Sum the columns using the addmargins function to answer the question of how many classmates were born in each season? Include your answer in the code.

```{r task6}
# Create bseason with mutate() and case_when(), following the same method used
# to categorize calories in the Starbucks demo. Because case_when() checks the
# conditions in order, months 1-2 become Winter, 3-5 become Spring, 6-8 become
# Summer, 9-11 become Fall, and month 12 becomes Winter.
C1survey <- C1survey %>%
  mutate(
    bseason = case_when(
      bmonth <= 2 ~ "Winter",
      bmonth <= 5 ~ "Spring",
      bmonth <= 8 ~ "Summer",
      bmonth <= 11 ~ "Fall",
      bmonth == 12 ~ "Winter"
    )
  )

# Convert bseason into a factor and label the levels in a logical order, using
# the factor() method demonstrated in the Starbucks demo.
C1survey$bseason <- factor(
  C1survey$bseason,
  levels = c("Winter", "Spring", "Summer", "Fall")
)

# Check that bseason is now a factor variable.
class(C1survey$bseason)

# Create a table with birth months in the rows and seasons in the columns.
# Include missing values so the missing birth month is not hidden.
season_table <- table(
  C1survey$bmonth,
  C1survey$bseason,
  useNA = "always"
)
season_table

# Use the addmargins() function required by the question to sum the columns.
addmargins(season_table, margin = 1)

# Answer: 5 classmates were born in Winter, 5 in Spring, 5 in Summer,
# and 15 in Fall. One classmate has a missing birth month and season.
```


##### 7. Pick your favorite variable or variables to analyze, come up with a question you want to answer with that variable or variables, generate the code, and provide an answer to your question. Describe what you did using comments (i.e. #'s).

```{r task7}
# Question: What is the most common favorite season among classmates?
# Use table(), as demonstrated in the Starbucks demo, to count the answers.
table(C1survey$fav_season, useNA = "always")

# Answer: Fall is the most common favorite season. Seventeen classmates selected
# Fall, compared with 7 for Spring, 3 for Summer, and 4 for Winter.
```
